---
title: "Comparing cluster assignments"
bibliography: '`r system.file("REFERENCES.bib", package="hospitraceR")`'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    dpi = 320
)

library(hospitraceR)
library(hospitraceRVisualize)
library(ape)
library(tidyr)
library(ggplot2)
library(ggalign)
library(paletteer)

# An example dataset of isolates and their patient epidemiological metadata.
extdata_dir <- file.path("..", "..", "inst", "extdata")
load(file.path(extdata_dir, "example.RData"))

# Read in sequence file
dna_aln <- read.dna(file.path(extdata_dir, "example.fasta"), format = "fasta")

# Get all variable positions in the alignment
var_pos <- apply(dna_aln, 2, function(x) sum(x == x[1]) < nrow(dna_aln))

# Only keep variable positions
dna_var <- dna_aln[, var_pos]

# Get pairwise distances
snp_dist <- get_snp_dist_matrix(dna_var)

# Get the phylogenetic tree
phylo_tree <- get_phylo_tree(dna_var, snp_dist, "pars")

# Generate the cluster assignments using the SNP threshold and threshold-free methods
clusters_snp <- get_tn_clusters_snp_thresh(snp_dist, 10)
clusters_sv <- get_tn_clusters_sv_index(dna_var, snp_dist, adm_seqs, adm_pos_pt_seqs, dna_pt_labels, dates, phylo_tree)
```

This article goes into more detail on comparing different transmission cluster assignments. Since `hospitraceR` provides different methods for generating transmission clusters, one of the natural questions is to try and compare the clusters generated by these different methods. In some cases, we might want to compare the clusters generated by slightly different subsets of the data as well. For these purposes, `hospitraceR` provides a few complementary ways of comparing two cluster assignments:

- a **visual** comparison of the isolates shared between clusters, via `hospitraceRVisualize::plot_jaccard_similarity_heatmap()`;
- two **partition-similarity** statistics that summarise the agreement between two clusterings in a single number, the **Adjusted Rand Index** (ARI) and the **Adjusted Mutual Information** (AMI), both built on top of a contingency table of shared isolates; and
- an **epidemiological** concordance metric, the **fraction of converts with the same source** (FSS), which asks whether the two clusterings agree on *who infected whom*, not just on how the isolates are grouped.

Throughout, we compare the two clusterings produced in the [introductory article](hospitraceR.html): `clusters_snp`, from the hard 10-SNP threshold method, and `clusters_sv`, from the threshold-free shared-variant method of @Hawken2022.

# Visualizing the fraction of shared isolates

One of the simplest ways to compare two cluster assignments is to visualize the fraction of isolates that are shared between the two clusters. This can be done using the `hospitraceRVisualize::plot_jaccard_similarity_heatmap()` function. This function takes in two cluster assignments and returns a heatmap showing the overlap between the clusters by indicating the proportion of isolates in each cluster that are also present in the clusters generated by the other method. Specifically, it calculates the Jaccard similarity index for each pair of clusters, which is defined as the size of the intersection of the two clusters divided by the size of the union of the two clusters. We can calculate the Jaccard similarity index for each pair of clusters as:

$$J(A, B) = \frac{|A \cap B|}{|A \cup B|}$$

where $A$ and $B$ are the two clusters, $|A \cap B|$ is the size of the intersection of the two clusters, and $|A \cup B|$ is the size of the union of the two clusters. What does this tell us? It tells us the proportion of isolates in cluster $A$ that are also present in cluster $B$. For comparing cluster assignments, we can look at all such pairs of clusters and visualize the similarity between them.

Before we can use this function, we remove any singleton clusters, which are clusters that contain only one isolate – these are not very useful for comparison. We keep the singleton-removed clusters in new variables so that the full assignments remain available for the metrics computed later in this article:

```{r compare_clusters, fig.width=10, fig.height=10}
# remove singleton clusters from both clustering methods (for plotting only)
clusters_snp_shared <- remove_singleton_clusters(clusters_snp)
clusters_sv_shared <- remove_singleton_clusters(clusters_sv)

# compare the clusters
plot_jaccard_similarity_heatmap(clusters_snp_shared, clusters_sv_shared) +
    # change the color palette
    scale_fill_paletteer_c("grDevices::Mint", direction = -1) +
    # add a title
    ggtitle("Comparison of clusters using SNP threshold and threshold-free methods") +
    # add x and y axis labels
    xlab("Threshold-free clusters") +
    ylab("SNP threshold clusters") +
    # center the title and rotate x axis labels
    theme(
        plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1)
    )
```

The heatmap is a useful overview, but it does not give us a single number we can track as we change a clustering parameter or swap in a different subset of the data. For that, we turn to the contingency table and the summary statistics built on top of it.

# The contingency table of shared isolates

Both the Jaccard heatmap and the summary statistics below are built from the same object: a **contingency table** that cross-tabulates the two cluster assignments. Each cell counts how many isolates fall into a given cluster under the first assignment *and* a given cluster under the second. `cluster_contingency_table()` computes it directly. Only isolates present in both assignments are counted; if the two assignments are not over exactly the same set of isolates, the function warns and restricts itself to the common isolates.

```{r contingency}
cont_table <- cluster_contingency_table(clusters_snp, clusters_sv)
dim(cont_table)
```

If the two clusterings agree more than random chance, we expect the table to be mostly sparse. We can subset the first few rows and columns to get a sense of the structure:

```{r contingency_corner}
cont_table_shared <- cluster_contingency_table(clusters_snp_shared, clusters_sv_shared)
cont_table_shared[1:8, 1:8]
```

The table thus gives us a base from which to base our comparisons. We've already seen one way to visualize this, using a heatmap. But we can also perform certain operations on this table to get a single number to indicate the concordance between the two clusterings. The two most common metrics are the **Adjusted Rand Index** and the **Adjusted Mutual Information**.

# Adjusted Rand Index (ARI)

The **Adjusted Rand Index** measures how often the two clusterings agree on whether a *pair* of isolates belongs together, corrected for the agreement we would expect from two random partitions of the same sizes (Hubert and Arabie, 1985). It ranges from $-1$ to $1$: a value of $1$ means the two clusterings are identical, $0$ means they agree no more than chance, and negative values mean they agree *less* than chance.

There are two functions which calculate the ARI from different inputs. `adjusted_rand_index()` takes the two cluster assignments directly:

```{r ari}
adjusted_rand_index(clusters_snp, clusters_sv)
```

If you already have a contingency table — for example because you are computing several metrics from it, or sweeping over a parameter and want to avoid rebuilding the table each time — you can call `ari_from_contingency()` on the table instead.

# Adjusted Mutual Information (AMI)

The **Adjusted Mutual Information** measures agreement from an information-theoretic angle: how much knowing an isolate's cluster under one assignment tells you about its cluster under the other, again corrected for chance (Vinh, Epps and Bailey, 2010). It ranges from $0$ (the two clusterings are independent) to $1$ (they are identical). Where ARI counts agreement over *pairs* of isolates, AMI is sensitive to how the isolates are distributed across clusters of different sizes, so the two metrics need not move together.

As with ARI, there is a convenience wrapper that takes the assignments directly and a variant (`ami_from_contingency()`) that works from a precomputed contingency table.

```{r ami}
adjusted_mutual_information(clusters_snp, clusters_sv)
```

Both ARI and AMI here are well above what we would expect by chance, confirming the visual impression from the heatmap: the two methods agree on many more isolates than by chance. However, they're not close to 1, which means there are still significant composition differences that may be interesting.

# Epidemiological concordance: fraction of converts with the same source

ARI and AMI are purely structural, and compare partitions without any knowledge of the underlying epidemiology. But when studying transmission, the question we actually care about is often not "are the clusters the same shape?" but "do the two methods agree on *who acquired the organism from whom?*". The **fraction of converts with the same source** (FSS) answers exactly this, by using the epidemiological information that is often available to us in transmission studies.

A *convert* (also referred to as an *acquisition*) is a patient who is inferred to have acquired the organism during their stay (rather than carrying it on admission). The *source* of a convert is an admission-positive patient to whom the convert's acquisition can ultimately be traced back. `hospitraceR` derives these roles from each clustering via `cluster_patient_categorization()`, using admission status, surveillance cultures and collection dates. `fraction_convert_same_source()` asks the question: for the set of common converts between two clusterings, how many of them have the same inferred source in both clusterings?

```{r fss}
fss <- fraction_convert_same_source(
    clusters_snp,
    clusters_sv,
    dna_aln,
    dna_pt_labels,
    adm_seqs,
    dates,
    surv_df,
    converts_without_assigned_source = TRUE
)
fss
```

So even where the two methods carve up the isolates slightly differently, they agree on the inferred source for a large majority of converts.

There can be cases where the two clustering methods are built using slightly different epidemiological inputs. In such cases, the convenience wrapper `fraction_convert_same_source()` is not appropriate because it would categorize both clusterings against a single set of epidemiological inputs. Instead, you should build an isolate lookup for each clustering using `get_isolate_lookup()` and then pass the two lookups to `fraction_convert_same_source_from_lookups()`.

# Putting it together: sweeping across SNP thresholds

The summary statistics come into their own when we want to compare *many* clusterings at once.

At each threshold we build the contingency table once and reuse it for ARI and AMI, which is why the `*_from_contingency()` entry points are convenient here:

```{r sweep}
thresholds <- c(2, 5, 10, 15, 20, 25)

sweep_df <- do.call(rbind, lapply(thresholds, function(thr) {
    clusters_thr <- get_tn_clusters_snp_thresh(snp_dist, thr)

    # build the contingency table once, reuse for ARI and AMI
    ct <- cluster_contingency_table(clusters_thr, clusters_sv)

    fss <- fraction_convert_same_source(
        clusters_thr,
        clusters_sv,
        dna_aln,
        dna_pt_labels,
        adm_seqs,
        dates,
        surv_df,
        converts_without_assigned_source = TRUE
    )

    data.frame(
        threshold = thr,
        ARI = ari_from_contingency(ct),
        AMI = ami_from_contingency(ct),
        FSS = fss
    )
}))

sweep_df
```

To plot all three metrics together, we reshape the results into long format and draw one line per metric:

```{r sweep_plot, fig.width=8, fig.height=5}
sweep_long <- pivot_longer(
    sweep_df,
    cols = c(ARI, AMI, FSS),
    names_to = "metric",
    values_to = "value"
)

ggplot(sweep_long, aes(x = threshold, y = value, color = metric)) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    scale_color_paletteer_d("ggsci::lanonc_lancet") +
    labs(
        x = "SNP distance threshold",
        y = "Agreement with threshold-free clustering",
        color = "Metric",
        title = "Hard SNP threshold vs. threshold-free clustering"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
```

# References

<div id="refs"></div>
