---
title: "Epidemiological overlap analysis"
bibliography: '`r system.file("REFERENCES.bib", package="hospitraceR")`'
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    dpi = 320
)
```

A genomic cluster tells us that a group of isolates are closely related, but on its own it does not say *how* the organism moved between the patients in that cluster. `hospitraceR` answers that question by layering the patients' movements through the facility on top of the genomic clusters: it asks whether the patients in a cluster were in the same place at overlapping times, and uses that to categorize the likely transmission route. This article walks through that workflow, from a single isolate lookup table to a permutation test of whether the observed overlap exceeds chance.

We start, as in the [introductory article](hospitraceR.html), by clustering the example isolates with the threshold-free shared-variant method of @Hawken2022. Visualization is provided by the companion package [`hospitraceRVisualize`](https://github.com/theabhirath/hospitraceRVisualize).

```{r setup, message = FALSE}
library(hospitraceR)
library(hospitraceRVisualize)
library(ape)
library(ggplot2)

# An example dataset of isolates and their patient epidemiological metadata.
extdata_dir <- file.path("..", "..", "inst", "extdata")
load(file.path(extdata_dir, "example.RData"))
dna_aln <- read.dna(file.path(extdata_dir, "example.fasta"), format = "fasta")

# Keep isolates whose patients appear in the location traces, then the variable sites
dna_aln <- dna_aln[dna_pt_labels[labels(dna_aln)] %in% rownames(facility_trace), ]
dna_var <- dna_aln[, apply(dna_aln, 2, function(x) sum(x == x[1]) < nrow(dna_aln))]

snp_dist <- get_snp_dist_matrix(dna_var)
tree <- get_phylo_tree(dna_var, snp_dist, "pars")

clusters <- get_tn_clusters_sv_index(
    dna_var, snp_dist, adm_seqs, adm_pos_pt_seqs, dna_pt_labels, dates, tree
)
```

# The isolate lookup table

Almost every function below consumes the same table: a per-isolate lookup that joins each isolate's cluster to its patient, collection date, admission status, and surveillance history. `get_isolate_lookup()` builds it. The two surveillance-derived columns are what make the epidemiological reasoning possible:

- `prev_surv` is the most recent surveillance date strictly before the isolate;
- `prev_surv_neg` is `TRUE` only when the patient had a prior surveillance *and none of the surveillances before this isolate were positive* — a genuine previous-negative to current-positive conversion. This tells us which isolates specifically were detected as part of acquisition events.

```{r lookup}
isolate_lookup <- get_isolate_lookup(
    clusters, dna_var, dna_pt_labels, adm_seqs, dates, surv_df
)
head(isolate_lookup)
```

# Categorizing patients by transmission role

Within each cluster, we attempt to categorize patients based on different roles: an *index* who carried the organism on admission, *converts* who acquired it during their stay, *weak indices* whose first surveillance was already positive, and so on. `cluster_patient_categorization()` assigns each patient in each cluster one of these roles from its admission status, surveillance cultures, and collection dates. `flatten_cluster_patient_categorization()` turns the nested result into a tidy data frame, one row per cluster-patient pair:

```{r patient_cats}
patient_cats <- cluster_patient_categorization(isolate_lookup, surv_df)
patient_cat_df <- flatten_cluster_patient_categorization(patient_cats)

table(patient_cat_df$category)
```

The converts are the patients we most want to explain: each one acquired the organism somewhere, and the location traces are how we look for that "somewhere".

# Overlap between isolates

The raw ingredient is the overlap between *pairs* of isolates. `isolate_isolate_overlap()` takes the lookup and a patient-by-date trace matrix (here `facility_trace`, marking which facility each patient was in on each day) and, for every ordered donor-recipient pair, counts the days on which the two patients were in the same place. The window it searches runs from the donor's previous surveillance date up to the recipient's collection date — the period during which the donor could plausibly have transmitted to the recipient before the recipient's positive isolate was collected.

```{r iso_overlap}
iso_overlap <- isolate_isolate_overlap(isolate_lookup, facility_trace)
head(iso_overlap)
```

This is *concurrent* overlap: the two patients were physically co-located at the same time. `isolate_isolate_sequential_overlap()` instead captures *indirect* overlap, where the recipient later occupies a location the donor had been in earlier — a possible environmental or intermediary route. Pairs with any concurrent co-location are excluded from the sequential count, so the two notions do not double-count the same contact.

# From isolate pairs to cluster explanations

`cluster_isolate_overlap()` collapses the pairwise overlaps to the cluster level: for each isolate in a multi-patient cluster, does *any* other isolate in the same cluster overlap with it? Admission-positive isolates are reported as `NA`, since they were not acquired in the facility and so need no in-facility explanation.

```{r cluster_overlap}
cluster_overlap <- cluster_isolate_overlap(isolate_lookup, iso_overlap)
```

`categorize_cluster_overlap()` then labels each cluster by how well its conversions are explained. A cluster where the index is admission-positive and every convert has an overlap is `patient-to-patient`; one where a convert is left unexplained is `missing-intermediate`; and there are further categories for weak indices, multiply-colonized indices, and clusters that the location data cannot account for. `flatten_cluster_overlap_categorization()` tidies the result:

```{r overlap_cats}
overlap_cats <- categorize_cluster_overlap(isolate_lookup, cluster_overlap, surv_df)
overlap_cat_df <- flatten_cluster_overlap_categorization(overlap_cats)

table(overlap_cat_df$category)
```

Most multi-patient clusters fall into the explained `patient-to-patient` category, which is the signal we would expect if co-location within the facility is genuinely driving transmission.

# Convert events with overlap

To summarize a whole clustering with a single number, `fraction_convert_events_with_overlap()` reduces the cluster-level overlaps to the fraction of *convert events* (previous-negative patients who then turned positive) that have an overlap explanation. It returns the per-cluster fractions, but also carries the per-cluster numerator (`n_overlap`) and denominator (`n_converts`) as attributes, so that clusters can be pooled by their number of converts rather than averaged as equal-weight fractions:

```{r convert_fraction}
convert_overlap <- fraction_convert_events_with_overlap(cluster_overlap, isolate_lookup)

# convert-weighted pooled fraction across all clusters
sum(attr(convert_overlap, "n_overlap")) / sum(attr(convert_overlap, "n_converts"))
```

# Is the overlap more than chance?

A high fraction is only meaningful if it is higher than we would get by shuffling patients between clusters at random. `cluster_overlap_perm_test()` runs that comparison. It computes the observed pooled overlap fraction at facility, floor, and room level (and their sequential variants), then repeatedly permutes the cluster assignments — preserving the index/convert structure — and recomputes the fraction each time to build a null distribution.

The expensive isolate-pair overlaps depend only on patient movements, not on cluster labels, so they are computed once and reused across permutations. We use a modest number of permutations here to keep the article quick; a real analysis would use more.

```{r perm_test, message = FALSE}
perm <- cluster_overlap_perm_test(
    clusters, dna_var, dna_pt_labels, adm_seqs, adm_pos_pt_seqs,
    dates, surv_df, facility_trace, floor_trace, room_trace,
    nperm = 200, num_cores = 2
)
```

The test returns, for the observed clustering and for every permutation, the number of converts with overlap (`n_overlap`) and the number of converts (`n_converts`), each summed over the tested clusters. Their ratio is the pooled overlap fraction per trace type. (A real analysis would run the same test across many sequence types and sum the counts over all of them before dividing.)

```{r perm_aggregate}
trace_types <- rownames(perm$observed)
nperm <- 200

# observed pooled fraction (and counts) per trace type
obs_df <- data.frame(
    trace_type = trace_types,
    overlap_fraction = perm$observed[, "n_overlap"] / perm$observed[, "n_converts"],
    perm$observed,
    row.names = NULL
)

# null distribution: pooled fraction per trace type, per permutation
perm_df <- data.frame(
    trace_type = rep(trace_types, each = nperm),
    overlap_fraction = as.vector(t(perm$permuted[, "n_overlap", ] / perm$permuted[, "n_converts", ]))
)

# one-sided p-value: how often the null meets or exceeds the observed fraction
pvals <- vapply(seq_along(trace_types), function(j) {
    null_vals <- perm_df$overlap_fraction[perm_df$trace_type == trace_types[j]]
    (sum(null_vals >= obs_df$overlap_fraction[j], na.rm = TRUE) + 1) / (nperm + 1)
}, numeric(1))

obs_df$p_value <- pvals
obs_df
```

`hospitraceRVisualize::plot_overlap_perm_test()` draws the core figure: a violin of the null distribution for each trace type, with the observed fraction as a red diamond and the observed counts labelled above. It expects the trace type to be a factor whose levels set the x-axis order, shared between the two data frames:

```{r perm_plot, fig.width = 8, fig.height = 5}
perm_df$trace_type <- factor(perm_df$trace_type, levels = trace_types)
obs_df$trace_type <- factor(obs_df$trace_type, levels = trace_types)

# annotate each trace type with its permutation p-value
annot <- data.frame(
    trace_type = obs_df$trace_type,
    label = sprintf("p = %.3f", obs_df$p_value),
    y = 1.04
)

plot_overlap_perm_test(
    perm_df, obs_df,
    title = "Convert overlap vs. permuted null",
    subtitle = sprintf("%d clusters, %d permutations", length(perm$valid_clusters), nperm)
) +
    geom_text(data = annot, aes(x = trace_type, y = y, label = label), size = 3, inherit.aes = FALSE)
```

The pattern is clear and biologically sensible. The concurrent co-location fractions (facility, floor, room) sit well above their null distributions: converts really were co-located with a plausible donor far more often than random cluster assignment would produce. The sequential (indirect) fractions, by contrast, fall within or below their nulls, so once direct co-location is accounted for, indirect location sharing adds no signal beyond chance. The facility level shows the strongest effect, as expected, since being on the same floor or in the same room implies being in the same facility.

# References

<div id="refs"></div>
